---
title: "PCAWG fits"
editor_options: 
  chunk_output_type: console
output:   
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    fig_width: 20
    fig_height: 13
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set( warning=FALSE, message=FALSE)
knitr::opts_knit$set(width = 75)
```

We plot CCF histograms with annotations of every subclonal driver mutation; we omit plotting clonal river mutations. This markodwn re-creates:

- the subset of cases shown in Figure 1;
- all the remaining plots (Supplementary Data). 

To make these plots we read in the MAF ccf file processed in the [data markdown](data.html).

```{r}
library(data.table)
library(tidyverse)
library(ggrepel)
library(cowplot)
library(gtools)

maf_ccf <- fread("processed_data/maf_subclone_ccf.csv.gz")

maf_ccf <- maf_ccf %>%
  mutate(histology_abbreviation = str_remove_all(histology_abbreviation, ".clearcell"))
```

# CCF histograms {.tabset .tabset-fade .tabset-pills}

```{r, results="asis", fig.width = 20, fig.height = 13}
sample_to_idx <-
  distinct(maf_ccf, Tumor_Sample_Barcode, histology_abbreviation) %>%
  arrange(histology_abbreviation) %>%
  mutate(idx = paste0(round(row_number() / 100) * 100))
sample_ids_chunk <-
  split(sample_to_idx$Tumor_Sample_Barcode, sample_to_idx$idx)

for (ct in mixedsort(names(sample_ids_chunk))) {
  cat(paste0('## ', as.numeric(ct) + 1, '-', as.numeric(ct) + 100, ' \n'))
  samples <- sample_ids_chunk[[ct]]
  
  dfdrivers <- maf_ccf %>%
    filter(issubclonaldriver == TRUE) %>%
    filter(Tumor_Sample_Barcode %in% samples)
  
  if (dim(dfdrivers)[1] > 0) {
    g <- maf_ccf %>%
      filter(Tumor_Sample_Barcode %in% samples) %>%
      ggplot(aes(x = ccf, fill = cluster)) +
      geom_histogram(bins = 100,
                     position = "stack",
                     alpha = 0.6) +
      facet_wrap(histology_abbreviation ~ icgc_sample_id, scales = "free_y") +
      geom_text_repel(
        data = dfdrivers,
        aes(label = gene, y = Inf, x = ccf),
        col = "black",
        alpha = 0.85,
        size = 2.0
      ) +
      geom_vline(
        data = dfdrivers,
        aes(xintercept = ccf, col = cluster),
        lty = 2,
        size = 0.25,
        show.legend = FALSE
      ) +
      scale_color_manual(values = RColorBrewer::brewer.pal(6, "Dark2"),
                         breaks = paste0("cluster", 1:6)) +
      scale_fill_manual(values = RColorBrewer::brewer.pal(6, "Dark2"),
                        breaks = paste0("cluster", 1:6)) +
      theme_cowplot(font_size = 11) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) +
      scale_x_continuous(breaks = c(0.0, 1.0, 2.0), limits = c(0, 2)) +
      panel_border() +
      xlab("CCF") +
      ylab("Counts")
    
    print(g)
  } 
  else 
  {
    g <- maf_ccf %>%
      filter(Tumor_Sample_Barcode %in% samples) %>%
      ggplot(aes(x = ccf, fill = cluster)) +
      geom_histogram(bins = 100,
                     position = "stack",
                     alpha = 0.6) +
      facet_wrap(histology_abbreviation ~ icgc_sample_id, scales = "free_y") +
      scale_color_manual(values = RColorBrewer::brewer.pal(6, "Dark2"),
                         breaks = paste0("cluster", 1:6)) +
      scale_fill_manual(values = RColorBrewer::brewer.pal(6, "Dark2"),
                        breaks = paste0("cluster", 1:6)) +
      theme_cowplot(font_size = 11) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) +
      scale_x_continuous(breaks = c(0.0, 1.0, 2.0), limits = c(0, 2)) +
      panel_border() +
      xlab("CCF") +
      ylab("Counts")
    
    print(g)
  }
  
  cat(' \n \n')
}
```

```{r, echo = FALSE}
sample_to_idx <-
  distinct(maf_ccf, Tumor_Sample_Barcode, histology_abbreviation) %>%
  arrange(histology_abbreviation) %>%
  mutate(idx = paste0(round(row_number() / 100) * 100))
sample_ids_chunk <-
  split(sample_to_idx$Tumor_Sample_Barcode, sample_to_idx$idx)

# PDF output
pdf("FigureS1.pdf", width = 20, height = 13)

for (ct in mixedsort(names(sample_ids_chunk))) 
  {
  samples <- sample_ids_chunk[[ct]]
  
  dfdrivers <- maf_ccf %>%
    filter(issubclonaldriver == TRUE) %>%
    filter(Tumor_Sample_Barcode %in% samples)
  
  if (dim(dfdrivers)[1] > 0) {
    g <- maf_ccf %>%
      filter(Tumor_Sample_Barcode %in% samples) %>%
      ggplot(aes(x = ccf, fill = cluster)) +
      geom_histogram(bins = 100,
                     position = "stack",
                     alpha = 0.6) +
      facet_wrap(histology_abbreviation ~ icgc_sample_id, scales = "free_y") +
      geom_text_repel(
        data = dfdrivers,
        aes(label = gene, y = Inf, x = ccf),
        col = "black",
        alpha = 0.85,
        size = 2.0
      ) +
      geom_vline(
        data = dfdrivers,
        aes(xintercept = ccf, col = cluster),
        lty = 2,
        size = 0.25,
        show.legend = FALSE
      ) +
      scale_color_manual(values = RColorBrewer::brewer.pal(6, "Dark2"),
                         breaks = paste0("cluster", 1:6)) +
      scale_fill_manual(values = RColorBrewer::brewer.pal(6, "Dark2"),
                        breaks = paste0("cluster", 1:6)) +
      theme_cowplot(font_size = 11) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) +
      scale_x_continuous(breaks = c(0.0, 1.0, 2.0), limits = c(0, 2)) +
      panel_border() +
      xlab("CCF") +
      ylab("Counts")
    
    print(g)
  } 
  else 
  {
    g <- maf_ccf %>%
      filter(Tumor_Sample_Barcode %in% samples) %>%
      ggplot(aes(x = ccf, fill = cluster)) +
      geom_histogram(bins = 100,
                     position = "stack",
                     alpha = 0.6) +
      facet_wrap(histology_abbreviation ~ icgc_sample_id, scales = "free_y") +
      scale_color_manual(values = RColorBrewer::brewer.pal(6, "Dark2"),
                         breaks = paste0("cluster", 1:6)) +
      scale_fill_manual(values = RColorBrewer::brewer.pal(6, "Dark2"),
                        breaks = paste0("cluster", 1:6)) +
      theme_cowplot(font_size = 11) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) +
      scale_x_continuous(breaks = c(0.0, 1.0, 2.0), limits = c(0, 2)) +
      #ggtitle(ct) +
      panel_border() +
      xlab("CCF") +
      ylab("Counts")
    
    print(g)
  }
}
dev.off()
```

# Figure 1

The sample ids for the 41 cases we include in Figure 1 of our letter are available as file `PCAWG_subclonal_drivers.csv`. 

```{r}
library(kableExtra)

readr::read_csv("PCAWG_subclonal_drivers.csv") %>% 
  filter(include == "Y")
  kbl() %>%
  kable_styling()
```


We now pull these cases out and plot the CCF histograms, making figure 1 in our letter.

```{r figure1, fig.cap="Figure 1: 41 cases with subclonal drivers"}
md <- fread("metadata/PCAWG_subclonal_drivers.csv") %>% 
   filter(include == "Y")

samples_to_include <- md$Tumor_Sample_Barcode

dfdrivers <- maf_ccf %>% 
  mutate(cluster = str_remove(cluster, "cluster")) %>% 
  filter(issubclonaldriver == TRUE) %>% 
  filter(Tumor_Sample_Barcode %in% samples_to_include) %>% 
  arrange(Tumor_Sample_Barcode, ccf) %>% 
  group_by(Tumor_Sample_Barcode) %>% 
  mutate(x = row_number()) %>% 
  mutate(x = ifelse(x == 1, max(ccf - 0.4, 0.0), ccf + 0.3))

g <- maf_ccf %>%
    mutate(cluster = str_remove(cluster, "cluster")) %>% 
    filter(Tumor_Sample_Barcode %in% samples_to_include) %>% 
    ggplot(aes(x = ccf, fill = cluster)) +
    geom_histogram(bins = 100, position = "stack", alpha = 0.6, size = 0.0) +
    facet_wrap(~icgc_sample_id, scales = "free_y") +
    geom_text_repel(data = dfdrivers,
                    force = 1,
                    aes(label = gene, y = Inf, x = x), col = "black",
                    alpha = 0.85, size = 1.5) +
    geom_vline(data=dfdrivers,
               aes(xintercept = ccf, col = cluster), lty = 2, size = 0.25, show.legend = FALSE) +
    scale_color_manual(values = RColorBrewer::brewer.pal(6, "Dark2"), breaks = paste0(1:6)) +
    scale_fill_manual(values = RColorBrewer::brewer.pal(6, "Dark2"), breaks = paste0(1:6)) +
    theme_cowplot(font_size = 8, line_size = 0.25) +
    theme(legend.position = c(0.9, 0.075)) +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) +
    scale_x_continuous(breaks = c(0.0, 1.0, 2.0), limits = c(0,2)) +
    labs(fill = "Cluster") +
    guides(fill=guide_legend(nrow=2)) +
    panel_border() +
    xlab("CCF") +
    ylab("Counts")
g
```


```{r}
save_plot(filename = "Figure1.png", plot = g, base_height = 5, base_width = 7)
save_plot(filename = "Figure1.pdf", plot = g, base_height = 5, base_width = 7)
```

For these 41 cases we also pull out  clustering probabilities for each cluster, which are used in one of our supplementary figures.

```{r}
library(glue)

dfclust <- data.frame()

for (sa in samples_to_include) {
  clust_assignments <-
    fread(glue("subclones_data/{sa}_cluster_assignments.txt.gz"))[mut_type != "SV"] %>%
    mutate(sample = sa) %>%
    select(sample, everything(),-chromosome2,-position2)
  dfclust <- bind_rows(dfclust, clust_assignments)
}

fwrite(dfclust, "processed_data/cluster_probs_41cases.csv.gz")
```
